Experiences with Planning for Natural Language Generation

### 2008-06-23 trunk/Papers/experiences/discussion.tex
changes to experiments (GIVE subsection)

The positive conclusion we can draw from the experiments reported above is
that modern planners are fairly good at controlling the search for many of
the moderately-sized NLG problems we looked at. This is particularly true
for FF in the sentence generation domain, which generates nontrivial
14-word sentences in about two seconds. There is still room for
improvement, however. For instance, by comparison a special-purpose
algorithm for referring expressions could generate ten of these 14-word
sentences in a few milliseconds. Although search efficiency is not the main
performance metric in these domains, it nevertheless becomes a factor in
scenarios like Experiment~2, where the world is almost completely
unconstrained and (unsurprisingly) the search becomes much slower.

>>>>

The positive conclusion we can draw from the simple experiments reported
above is that modern planners are fairly good at controlling the search for
many of the moderately-sized NLG problems we looked at. This is
particularly true for FF in the sentence generation domain, which generates
nontrivial 14-word sentences in about two seconds. There is still room for
improvement, however. For instance, by comparison a special-purpose
algorithm for referring expressions could generate ten of these 14-word
sentences in a few milliseconds. Although search efficiency is not the main
performance metric in these domains, it nevertheless becomes a factor in
scenarios like Experiment~2, where the world is almost completely
unconstrained and (unsurprisingly) the search becomes much slower.

### 2008-06-23 trunk/Papers/experiences/experiments.tex
changes to experiments (GIVE subsection)

We now describe the results of a set of experiments designed to evaluate
the suitability of particular planners (FF, SGPLAN, and an ad-hoc
implementation of GraphPlan) in our two NLG domains. As a first impression,
our results indicate that planning is a promising tool for both domains. In
the GIVE domain, SGPLAN 5.2.2 computes a domain plan from the initial state
to the goal in 0.3 seconds, which is fast enough for moderately-sized
problems instances in the application.\footnote{All runtimes were
  measured on a Pentium 4 CPU running at 3 GHz. Java programs were allowed
  to ``warm up'', i.e.\ the planner was run five times and the first four
  measurements discarded to ensure that the JVM had just-in-time compiled
  all relevant bytecode.}
In the sentence generation domain, FF dramatically outperforms the best
previously known algorithm for the same problem (a reimplementation of
\cite{Stone2003a}), although the latter is a greedy search algorithm with a
heuristic that is hand-tailored to the domain.

@@

In the second experiment, we evaluate the performance of the planners on
problems arising in the GIVE domain. We construct a series of test worlds,
similar to the one illustrated in Figure~\ref{fig:give-minimal}. These worlds
consist of a $2n$ by $h$ grid of positions, such that there are buttons at
positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i \leq n$. The player starts
in position $(1,1)$ and must press all buttons in order to complete the
game. The world is generated as a GIVE world description, and then
automatically converted into a planning problem by the GIVE software.  (We
refer the reader to \todo{cite?} for more details about this process.)

@@

Some results from this experiment (for $h=20$, with $n$ ranging from
$1$ to $40$) are shown in Figure~\ref{fig:give-runtime-minimal}.  The
most obvious result is that FF is unable to solve any problems beyond
$n=13$ on the experimentation machine within the memory limit of 1 GB.
SGPLAN, on the other hand, solves instances beyond $n=40$ without
major problems.  The time spent on grounding is not a major factor in
either planner, probably because the planners need more time to
actually compute the plan -- at $n=40$, the optimal plan has a length
of about 1600!

@@

Finally, we varied the GIVE world in order to judge the effect of the
universe size in the GIVE domain (see Figure~\ref{fig:give-junk}).  We
added another $w$ by $h$ positions to the right of a minimal world as
in Experiment 2.  Other than that, we left the initial state and goal
untouched, i.e.\ the new positions were not needed in any plan; this
approximates the situation in the actual GIVE domain, where most grid
positions never need to be used.  As before, we generate a GIVE world
description and then convert it into a planning problem.

\begin{figure}
  \centering
  \includegraphics[width=1\columnwidth]{pic-empty-buttons}
  \caption{GIVE world with extra positions.}
  \label{fig:give-junk}
\end{figure}

Some results for $h=20$ and $n=5$ (with $w$ ranging from $1$ to $70$)
are shown in Figure~\ref{fig:give-runtime-junk}.  Again, FF runs out of
memory, this time at $w=17$, whereas SGPLAN happily solved inputs
beyond $w=70$.  However, unlike in Experiment 2, both planners now
spend a substantial propertion of their time on grounding again.  In
SGPLAN, this translates to a ``parsing time'' (which
\todo{presumably} includes grounding) which grows from 180 ms to 21.7
seconds as $w$ grows from $1$ to $75$; the rest of the runtime (which
includes the search) only grows from 400 ms to 2.3 seconds.  This
difference is particularly dramatic given that the actual optimal plan
is an identical plan of about 100 steps for each $w$.

>>>>

We now describe the results of three simple experiments designed to
evaluate the suitability of particular planners (FF, SGPLAN, and an ad-hoc
implementation of GraphPlan) in our two NLG domains. As a first impression,
our results indicate that planning is a promising tool for both domains. In
the GIVE domain, SGPLAN 5.2.2 computes a domain plan from the initial state
to the goal in 0.3 seconds, which is fast enough for moderately-sized
problems instances in the application.\footnote{All runtimes were
  measured on a Pentium 4 CPU running at 3 GHz. Java programs were allowed
  to ``warm up'', i.e.\ the planner was run five times and the first four
  measurements discarded to ensure that the JVM had just-in-time compiled
  all relevant bytecode.}
In the sentence generation domain, FF dramatically outperforms the best
previously known algorithm for the same problem (a reimplementation of
\cite{Stone2003a}), although the latter is a greedy search algorithm with a
heuristic that is hand-tailored to the domain.

@@

In the second experiment, we evaluate the performance of the planners on
problems arising in the GIVE domain. We construct a series of test worlds,
similar to the one illustrated in Figure~\ref{fig:give-minimal}. These
worlds consist of a $2n$ by $h$ grid of positions, such that there are
buttons at positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i \leq n$. The
player starts in position $(1,1)$ and must press all buttons in order to
successfully complete the game. The world is generated as a GIVE world
description, and then automatically converted into a planning problem by
the GIVE software. (We refer the reader to \todo{add citation} for more
details about this process.)

@@

Results for the $h=20$ case, with $n$ ranging from $1$ to $40$, are shown
in Figure~\ref{fig:give-runtime-minimal}.  The most obvious result is that
FF is unable to solve any problems beyond $n=13$ on our experimentation
machine within the memory limit of 1 GB.  SGPLAN, on the other hand, solves
instances beyond $n=40$ without major problems.  The time spent on
grounding is not a major factor in either planner, probably because the
planners need more time to actually compute the plan---for instance, the
optimal plan for the problem $n=40$ has a length of about 1600 steps.

@@

In the final experiment, we vary the structure of the GIVE world in order
to judge the effect that universe size has on the planning problem in this
domain.  Starting with the ordinary GIVE world described in Experiment~2,
we add another $w$ by $h$ empty ``junk'' positions to the right of the
minimal world (see Figure~\ref{fig:give-junk}). These new positions are not
actually needed in any plan, but approximate the situation in the actual
GIVE domain, where most grid positions never need to be used. We leave the
initial state and goal untouched. As before, we generate a GIVE world
description and then convert it into a planning problem.

\begin{figure}
  \centering
  \includegraphics[width=1\columnwidth]{pic-empty-buttons}
  \caption{GIVE world with extra ``junk'' positions.}
  \label{fig:give-junk}
\end{figure}

Results for the $h=20$, $n=5$ case (with $w$ ranging from $1$ to $70$) are
shown in Figure~\ref{fig:give-runtime-junk}. As in Experiment~2, FF again
runs out of memory, this time at $w=17$. SGPLAN happily solves inputs
beyond $w=70$. However, unlike in Experiment 2, both planners now spend a
substantial propertion of their time on grounding. In SGPLAN, this
translates to a ``parsing time'' (which \todo{presumably} includes
grounding) which grows from 180 ms to 21.7 seconds as $w$ grows from $1$ to
$75$. The rest of the runtime (which also includes the search time) only
grows from 400 ms to 2.3 seconds. This difference is particularly dramatic
given that the actual optimal plan in each case is an identical plan of
about 100 steps. The overall runtimes in these instances is also concerning
since runtimes over a couple seconds negatively affect the response time of
the GIVE system.
